Composition-climate paper

### 2016-08-24 Composition_Ecology.Rmd
Changes through the Results and into the Discussion.  Fixed `R/Setup_R.R` to move the nldc raster into `data/input`. This will allow complete reproduction of the paper from a git clone.
Removed the `gitignore` from the `data/albers/density` folder.

In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced by the land uses associated with EuroAmerican Settlement in the 1800s [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical] outbreaks of exotic pathogens [@anagnostakis1987chestnut;@barnes1979stem], and the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].  Modern forests, as represented by the Forest and Inventory Analysis surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@gorng2016preset;@schulte2007homogenization]. Wildland ecosystems have been widely converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f).  In northern forested regions, ranges of early successional species have expanded and abundances of late-successional species are reduced due to widespread logging in the 19^th^ and 20^th^ century, and ongoing land management [@schulte2007homogenization;@goring2016preset].  Additionally, formerly dominant tree species, *Ulmus americana* and *Castanea dentata* have experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem;@anagnostakis1987chestnut]. These vegetation changes are superimposed on strong regional gradients of temperature and precipitation in the upper Midwest and two major ecotones, one from prairie/savanna to closed forest, in an approximately west to east direction, and one from conifer dominated, sub-boreal forests to deciduous forest in a north to south direction.

Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historic climate and vegetation changes on these shifts.  To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@gorng2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.  We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and relative importance of historic vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.

@@

Estimates of settlement-era forest composition and tree distribution based on survey data from the Public Land Survey System [PLS: @white1983history;@schulte2001original;@gorng2016preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@gorng2016preset]. We briefly review the methods here. PLS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 - 1904 [@schulte2001original;@almendinger1996minnesota;@white1983history]. At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuths to each tree. Goring et al. [-@gorng2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@gorng2016preset]. Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (`r tab_nums("genera_table", display = "cite")`). A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus* to Ironwood, and *Thuja* and *Juniperus* to Cedar. 'Here 'Presence' is defined as at least one recorded tree in any 8x8 km grid cell.

@@

Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@gorng2016preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  For example, the FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data. This leads to higher cell-to-cell variability in the FIA data, but these are not expected to affect the regional-scale analyses here, particularly because we are investigating presence-absence patterns. Additionally, FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing the spatial extent of many tree taxa to be lower in the FIA than in the PLS datasets (`r tab_nums("genera_table", display="cite")`). Hence, scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA. However the practice of using FIA data as a surrogate for tree presence-absence data is widespread [@iverson2013tree;@iverson1998predicting], so this complication is a general one faced by forest ecologists and global change ecologists.  In prior comparative analyses of the PLS and FIA data, Goring et al. [-@gorng2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.

@@

This work reinforces the importance of historical datasets for examining the relationships between species and climate, and their importance for forecasting future species distributions [@moritz2013future;@maguire2015modeling]. Modern surveys sample only a subset of historical distributions, both spatially and within multidimensional climate space.  These analyses demonstrate that species-climate relationships have already begun to change, as a result of land use and climate chnage over the past few centuries, and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat]. Land use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000global]. While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@gorng2016preset;@kujawa2016theeffect], and the correlational structure between species and climate.

>>>>

In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced by the land uses associated with EuroAmerican Settlement in the 1800s [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical] outbreaks of exotic pathogens [@anagnostakis1987chestnut;@barnes1979stem], and the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].  Modern forests, as represented by the Forest and Inventory Analysis surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization]. Wildland ecosystems have been widely converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f).  In northern forested regions, ranges of early successional species have expanded and abundances of late-successional species are reduced due to widespread logging in the 19^th^ and 20^th^ century, and ongoing land management [@schulte2007homogenization;@goring2016preset].  Additionally, formerly dominant tree species, *Ulmus americana* and *Castanea dentata* have experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem;@anagnostakis1987chestnut]. These vegetation changes are superimposed on strong regional gradients of temperature and precipitation in the upper Midwest and two major ecotones, one from prairie/savanna to closed forest, in an approximately west to east direction, and one from conifer dominated, sub-boreal forests to deciduous forest in a north to south direction.

Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historic climate and vegetation changes on these shifts.  To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.  We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and relative importance of historic vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.

@@

Estimates of settlement-era forest composition and tree distribution based on survey data from the Public Land Survey System [PLS: @white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@goring2016preset]. We briefly review the methods here. PLS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 - 1904 [@schulte2001original;@almendinger1996minnesota;@white1983history]. At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuths to each tree. Goring et al. [-@goring2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset]. Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (`r tab_nums("genera_table", display = "cite")`). A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus* to Ironwood, and *Thuja* and *Juniperus* to Cedar. 'Here 'Presence' is defined as at least one recorded tree in any 8x8 km grid cell.

@@

Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@goring2016preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  For example, the FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data. This leads to higher cell-to-cell variability in the FIA data, but these are not expected to affect the regional-scale analyses here, particularly because we are investigating presence-absence patterns. Additionally, FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing the spatial extent of many tree taxa to be lower in the FIA than in the PLS datasets (`r tab_nums("genera_table", display="cite")`). Hence, scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA. However the practice of using FIA data as a surrogate for tree presence-absence data is widespread [@iverson2013tree;@iverson1998predicting], so this complication is a general one faced by forest ecologists and global change ecologists.  In prior comparative analyses of the PLS and FIA data, Goring et al. [-@goring2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.

@@

This work reinforces the importance of historical datasets for examining the relationships between species and climate, and their importance for forecasting future species distributions [@moritz2013future;@maguire2015modeling]. Modern surveys sample only a subset of historical distributions, both spatially and within multidimensional climate space.  These analyses demonstrate that species-climate relationships have already begun to change, as a result of land use and climate chnage over the past few centuries, and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat]. Land use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000global]. While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@goring2016preset;@kujawa2016theeffect], and the correlational structure between species and climate.
